---
title: "preliminary_model"
author: "Erin Xu & Dora Dong"
date: "2025-05-19"
output: html_document
---

```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("tidyverse", "latex2exp", "gridExtra", "rvest", "kableExtra")
lapply(packages, require, character.only = T)
set.seed(420)
```

# Load data and check if needed to clean. 
```{r, echo=FALSE}
all_data <- read.csv("water_pollution_disease.csv", na.strings = c("", "NA"))
anyNA(all_data)
sum(is.na(all_data))
colSums(is.na(all_data))
```

# Check for regression assumptions. 
```{r, echo=FALSE}
```
# clean data
``` {r echo=FALSE}
all_data <- read.csv("water_pollution_disease.csv", na.strings = c("", "NA"))
all_data <- na.omit(all_data) 
```

# Get the response and predictors. 
```{r echo=FALSE}
response <- all_data$Diarrheal.Cases.per.100.000.people
x0 <- all_data$Contaminant.Level..ppm.
x1 <- all_data$Bacteria.Count..CFU.mL.
x2 <- all_data$Sanitation.Coverage....of.Population.
x3 <- all_data$GDP.per.Capita..USD.
x4 <- all_data$Access.to.Clean.Water....of.Population.
x5 <- all_data$Healthcare.Access.Index..0.100.

all_data$Water_Treatment_Method <- as.factor(all_data$Water.Treatment.Method)
all_data$Water_Source_Type <- as.factor(all_data$Water.Source.Type)

model <- lm(response ~ x0 + x1 + x2 + x3 + x4 + x5 + Water_Treatment_Method + Water_Source_Type, data=all_data)
summary(model)
```

# Residual plots
```{r message=FALSE, echo=FALSE, warning=FALSE}
par(mfrow = c(2, 2))
plot(model)
```




################# Life_Expectancy_Data #############################
####################################################################
####################################################################


---
title: "preliminary_model"
author: "Erin Xu & Dora Dong"
date: "2025-05-19"
output: html_document
---
  
```{r message=FALSE, echo=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("tidyverse", "latex2exp", "gridExtra", "rvest", "kableExtra")
lapply(packages, require, character.only = TRUE)
set.seed(420)
```

# Load data and check if needed to clean. 
```{r, echo=FALSE}
all_data <- read.csv("Life_Expectancy_Data.csv", na.strings = c("", "NA"))
anyNA(all_data)
sum(is.na(all_data))
colSums(is.na(all_data))

# clean data
``` {r echo=FALSE}
all_data <- read.csv("Life_Expectancy_Data.csv", na.strings = c("", "NA"))
all_data <- na.omit(all_data) 
```

## Get the response and predictors. 

```{r echo=FALSE}
response <- all_data$Life.expectancy
x0 <- all_data$Adult.Mortality
x1 <- all_data$Schooling
x2 <- all_data$Income.composition.of.resources
x3 <- all_data$percentage.expenditure
x4 <- all_data$Diphtheria
x5 <- all_data$Polio
x6 <- all_data$Alcohol
x7 <- all_data$BMI

all_data$Status <- as.factor(all_data$Status)

model <- lm(response ~ x0 + x1 + x2 + x3 + x4 + x5 + x6 + x7 + Status, data=all_data)
summary(model)
```

# Residual plots
```{r,message=FALSE, echo=FALSE,warning=FALSE}
par(mfrow = c(2, 2))
plot(model)
```

















